Simulations of gravity-driven flow of binary liquids in microchannels

### 2011-03-29 Review/answers.tex
Started producing high quality vector plots.

>>>>

\textbf{Question:}
\begin{quotation}
Three ﬁgures in the manuscript are scanned from the litterature (Ref. 3, 10 and 16). This is
unusual
for an article which is not a review. I encourage the Authors to remove at least the two ﬁrst ones
which are not essential.
\end{quotation}

\textbf{Answer:} We eliminated necessary pictures and references to them.

\textbf{Question:} 
\begin{quotation}
In Figure 7, the points of data reported from the numerical work of Ref. 3 exhibit a surprisingly
noisy trend that corresponds more to an experimental study. The Authors should explain how they
obtain this data and plot them.
\end{quotation}

\textbf{Answer:} We added a description to the figure caption explaining data acquisition.

\textbf{Question:} 
\begin{quotation}
The plots of Figure 8 are of less quality than the other ﬁgures of the paper.
\end{quotation}

\textbf{Answer:} Unfortunately, it was not possible to extract data as a picture from the paper by
\citet{sehgal-microchannel}. Thus, the curves were extracted from the graphs and represented in a
new figure scaled to the present simulations for comparison reasons.

\textbf{Question:}
\begin{quotation}
The work of Yang seems to be close from the present study since it involves the
solution of Bretherton
ﬂow with a Lattice Boltzmann method. The Authors should discuss this paper and emphasize the
differences between their work and this paper.
\end{quotation}

\textbf{Answer:} We outlined the differences in the introduction section.

\textbf{Question:} It would be interesting if the Authors could mention a study in which a level-set
method is used in
order to complete the bibliography.

\textbf{Answer:} We added references to the introduction section.

\textbf{Questions:} 
\begin{quotation}
A secondary quantity of interest is the bubble speed or in a non-dimensional
form the capillary
number. This quantity is measured at the center of the bubbles. The Authors should describe how
this quantity as well as the center of the bubble are measured.
\end{quotation}

\textbf{Answer:} Thank you for pointing that. The classical Bretherton problem
$U_{\mathrm{bubble}}$ is the bubble velocity, i.e. the interface velocity. We recalculate all the
results taking into the consideration the interface velocity at the bubble tip. Qualitatively, the
results are the same. It was found that Poiseuille profile assumption is improved significantly
after recalculation.

\textbf{Question:} 
\begin{quotation}
 If possible, the Authors should present results showing that steady state is reached. It would
also
be interesting if the Authors show some patterns of the ﬂow by plotting for instance the vector ﬁeld
of velocity around the bubbles.
\end{quotation}

\textbf{Answer:} We added additional section which shows the change of flow pattern.

\textbf{Question:}
\begin{quotation}
In the Abstract, the Authors should give the range of real capillary numbers they measured instead
of the targeted values (Calit ).
\end{quotation}

\textbf{Answer:} We changed it to the measured capillary numbers.

\textbf{Question:} 
\begin{quotation}
The Authors should discuss more their results (Sections 5-4, 5-5, 5-6). Did the Authors measure the
pressure drops at the front and rear interfaces of the bubbles?
\end{quotation}

### 2011-03-29 Text/paper.tex
Started producing high quality vector plots.

\begin{abstract}
%This work shows that the free-energy binary liquid lattice Boltzmann method is able to
%reproduce the gravity driven microchannel flow of a gas bubble through liquid. In the regime of low
%capillary
%numbers
%$\Ca$ (defined in Eq.~\ref{capillary:number:definition}), the flow is thoroughly described by
%Bretherton \cite{bretherton},
%with the deposition film thickness proportional to $\Ca^{2/3}$. However, there are deviations
%from the Bretherton correlation for $\Ca\geq 0.005$, studied experimentally and numerically. This
%work aims to examine the applicability of the lattice Boltzmann binary liquid model to simulations
%of bubble flow in liquids for a range of $\Ca$ numbers. The binary
%liquid model is a continuous-interface multiphase model allowing to properly resolve the interface.
%It was found that the interface thickness needs to be $50$
%percent of the film thickness for the simulation to give convergent results. When
%the convergence criterion is fulfilled, the results are in a good agreement with the
%numerical simulations of \citet{giavedoni-numerical} in the range of $\Ca$ from $0.01$ to $1$. 
In this work a free-energy binary liquid lattice-Boltzmann scheme is used to simulate
Taylor/Bretherton flow in a micro-channel where elongated gas bubbles move through a liquid with
thin liquid films between the bubbles and the channel walls. The numerical scheme has a diffuse
interface, and a main focus of our work is to assess resolution requirements for correctly resolving
the liquid film and bubble motion. The simulations are two-dimensional and span a capillary number
range of $0.03$ to $0.8$ where the capillary number is based on the liquid’s dynamic viscosity, the
velocity of the bubble, and the interfacial tension. The flow is driven by a body force, and
periodic boundary conditions apply in the streamwise direction. We obtain grid independent results
as long as the liquid film thickness is at least two times the width of the diffuse interface, with
film thicknesses in accordance to literature results. We also show that the results in terms of film
thicknesses are largely insensitive to the liquid-gas viscosity ratio and wettability parameters. 
\end{abstract}

@@

One should acknowledge the works of \citet{pagonabarraga-fingers} on menisci
in thin films for fingering phenomena. \citet{sehgal-microchannel} performed lattice Boltzmann
simulations of two-dimensional channel flows for a relatively large capillary numbers, and
found discrepancies with the classical Bretherton theory, which
is limited to the low capillary number regime \cite{giavedoni-numerical}.

@@

film thickness is taken from {\color{green} Fig.~\ref{fig:heil:planar}},
  and for Reynolds number $40$ equals $0.06$.  

@@

\includegraphics[width=0.97\textwidth]{Figures/giavedoni_planar.eps}
\caption{\citet{giavedoni-numerical} gathered results across the
literature for different capillary numbers (reprinted from \cite{giavedoni-numerical}). Results are
scaled to the half-width of the
channel. $h_{\infty}$ is the film thickness of air finger at
the infinite distance from the front meniscus. \label{fig:giavedoni:planar}}
\includegraphics[width=0.97\textwidth]{Figures/heil-planar.eps}
\caption{\citet{heil-bretherton} performed simulations for different ranges of
capillary numbers and Reynolds numbers (reprinted from \cite{heil-bretherton}). Results
are scaled to the half-width of the channel. $h_{0}$ is the film thickness of air finger at
the infinite distance from the front meniscus.
\label{fig:heil:planar}}

@@

\subsection{The film variation over the bubble}
\label{section:film:thickness:variation}
Next, we investigated the variation of the film thickness over the bubble length.
The comparison for the thickness variation is presented in Fig.~\ref{fig:sehgal:bubble:length}.
\begin{figure}
\includegraphics[width=0.47\textwidth]{Figures/Bubble/bubble_sehgal.eps}\hfill
\includegraphics[width=0.47\textwidth]{Figures/Bubble/bubble_length.eps}\\
\caption{A qualitative comparison for the film thickness across
the bubble length. Left (courtesy of \citet{sehgal-microchannel}),
right (present simulations); $x$ is scaled to $H_{\mathrm{eff}}$
and increases in the flow direction.\label{fig:sehgal:bubble:length}}
\end{figure}
One can see a qualitative agreement, i.e. the thickness increases towards the front meniscus and
rapidly decreases towards the rear meniscus. This shape is sometimes referred as a "bullet"
shape.

\subsection{Capillary number region}
\label{section:capillary:range}
The purpose of this section is to validate the correlations of
\citet{giavedoni-numerical} in Fig. \ref{fig:giavedoni:planar} and
\citet{heil-bretherton} in Fig. \ref{fig:heil:planar} for a range of capillary
numbers. Because of limited computational resources, we skip the
small capillary numbers and make calculations for the range of capillary numbers $0.03-0.8$, which
is a computationally reasonable task.  To be consistent with the previous sections, we choose the grid to be
$202 \times 3001$. Then $5$ lattice Boltzmann units do not occupy more than $60$
percent of the effective film thickness. Because the simulation gets unstable with
smaller grids and larger gradients, all the capillary number simulations were
performed on the same grid. To properly initialize the body force, the
proportionality law was utilized. The forcing
$6 \cdot 10^{-6}/16$ was chosen to obtain the predicted capillary
number $0.05$.  
The pressure gradient can be obtained through the capillary number
ratio, Section \ref{section:init}:
\begin{equation}
\frac{\mathrm{d}P}{\mathrm{d} x}=6 \cdot 10^{-6}/16 \frac{\Ca_{\mathrm{lit}}}{0.05}
\end{equation}
The film thickness is initialized through the ratio of capillary numbers as well:
\begin{equation*}
12 \frac{\Ca_{\mathrm{lit}}}{0.05}
\end{equation*}
The results obtained after $2\cdot10^5$ steps are presented in Table
\ref{table:parameters:capillary:number}. 
{\color{red} We chose to measure the film thickness in the middle of the bubble. The
classical Bretherton formulation is for the measurement of the film thickness at ``infinity''.
However, \citet{cerro-bubble-train} indicate that for the gravity-driven bubble train flow in tubes
it is enough for bubble to have a length of two-three diameters to apply analysis for very long
bubbles. In comparison with the axisymmetric circular geometry, for three-dimensional square shaped
microchannels \citet{heil-threedim} indicate the measurement of the film thickness at $5.5$
channel heights from the front bubble tip. This distance was shown sufficient for the
film thickness to establish itself for the capillary number range $\Ca<4$.
\citet{giavedoni-numerical} showed that the film stabilizes at the distances as $2.6-4.0$ diameters
from the front tip depending on the Reynolds number.  We chose to measure the film thickness in the
middle of the bubble, which is located at least at the distance of $2.5-3$ channel heights from the
front bubble tip. Moreover, the film thickness examination, see Section
\ref{section:film:thickness:variation}, shows that for the capillary range of interest, i.e.
$0.05\leq \Ca \leq 1$, the standard deviation from the bubble middle film thickness is around $7$
percent for small capillary numbers ($\Ca=0.05$) and less than $1$ percent for larger ones ($\Ca
\geq 0.1$) at the distances ranging from one channel height from the rear tip upto one channel
height to the front tip. This certainly validates the point of choice to measure bubble thickness.}
One can see that the calculated
capillary numbers are overpredicted but the obtained thicknesses are overpredicted
as well due to the body force setting caused by the Poiseuille flow assumption. If one
needs to obtain the desired capillary number, the shooting method for the forcing is
necessary, starting with the Poiseuille pressure gradient force as the initial condition. We extracted
the data of 
\citet{giavedoni-numerical} and \citet{heil-bretherton} and compared them with our results (see 
Fig. \ref{fig:capillary:comparison}).
\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/Capillary/capillaries_comparison_real.eps}
\caption{A comparison between simulation results and results of
\citet{giavedoni-numerical} and \citet{heil-bretherton}. One can see a
reasonable agreement. The plots depict the film thickness as a function of the
the capillary number.\label{fig:capillary:comparison}}
\end{figure}

>>>>

\begin{abstract}
%This work shows that the free-energy binary liquid lattice Boltzmann method is able to
%reproduce the gravity driven microchannel flow of a gas bubble through liquid. In the regime of low
%capillary
%numbers
%$\Ca$ (defined in Eq.~\ref{capillary:number:definition}), the flow is thoroughly described by
%Bretherton \cite{bretherton},
%with the deposition film thickness proportional to $\Ca^{2/3}$. However, there are deviations
%from the Bretherton correlation for $\Ca\geq 0.005$, studied experimentally and numerically. This
%work aims to examine the applicability of the lattice Boltzmann binary liquid model to simulations
%of bubble flow in liquids for a range of $\Ca$ numbers. The binary
%liquid model is a continuous-interface multiphase model allowing to properly resolve the interface.
%It was found that the interface thickness needs to be $50$
%percent of the film thickness for the simulation to give convergent results. When
%the convergence criterion is fulfilled, the results are in a good agreement with the
%numerical simulations of \citet{giavedoni-numerical} in the range of $\Ca$ from $0.01$ to $1$. 
In this work a free-energy binary liquid lattice-Boltzmann scheme is used to simulate
Taylor/Bretherton flow in a micro-channel where elongated gas bubbles move through a liquid with
thin liquid films between the bubbles and the channel walls. The numerical scheme has a diffuse
interface, and a main focus of our work is to assess resolution requirements for correctly resolving
the liquid film and bubble motion. The simulations are two-dimensional and span a capillary number
range of $0.05$ to $1.0$ where the capillary number is based on the liquid’s dynamic viscosity, the
velocity of the bubble, and the interfacial tension. The flow is driven by a body force, and
periodic boundary conditions apply in the streamwise direction. We obtain grid independent results
as long as the liquid film thickness is at least two times the width of the diffuse interface, with
film thicknesses in accordance to literature results. We also show that the results in terms of film
thicknesses are largely insensitive to the liquid-gas viscosity ratio and wettability parameters. 
\end{abstract}

@@

One should acknowledge the works of \citet{pagonabarraga-fingers} on menisci
in thin films for fingering phenomena. \citet{sehgal-microchannel} performed lattice Boltzmann
simulations of two-dimensional channel flows for a relatively large capillary numbers, and
found discrepancies with the classical Bretherton theory, which
is limited to the low capillary number regime \cite{giavedoni-numerical}. Authors didn't take into
the account recent studies extending microchannel simulations for the capillary numbers
beyond the Bretherton regime. Thus, limiting comparison with other established CFD methods. Author
used the Shan-Chen model \cite{Shan-chen:extended} to simulate the multiphase behavior. It is also
the continuous interface model, but the study of grid dependence is absent.

{\color{red}
Level set method is represented by a number of works \cite{fukugata-levelset,
lakehal-micro, carlson-comparison}. While the bubble shapes and  were found in a good agreeement
with the experiments, the range of studied bubble lengths is confined to $2-3$ channels diameters.
Thus, the bubble radius changes significantly and cannot be used as classical Bretherton problem
benchmark. However, the works provide insight to the bubble formation and small bubbles flow in the
microchannels. Also, \citet{carlson-comparison} found a significant influence of the mutual bubbles
motion on the flow patterns.}

@@

film thickness is taken from correlations by \citet{heil-bretherton} 
%{\color{green}
%Fig.~\ref{fig:heil:planar}}
, and for Reynolds number $40$ equals $0.06$.  

@@

%\begin{figure}
%\includegraphics[width=0.97\textwidth]{Figures/giavedoni_planar.eps}
%\caption{\citet{giavedoni-numerical} gathered results across the
%literature for different capillary numbers (reprinted from \cite{giavedoni-numerical}). Results are
%scaled to the half-width of the
%channel. $h_{\infty}$ is the film thickness of air finger at
%the infinite distance from the front meniscus. \label{fig:giavedoni:planar}}
%\includegraphics[width=0.97\textwidth]{Figures/heil-planar.eps}
%\caption{\citet{heil-bretherton} performed simulations for different ranges of
%capillary numbers and Reynolds numbers (reprinted from \cite{heil-bretherton}). Results
%are scaled to the half-width of the channel. $h_{0}$ is the film thickness of air finger at
%the infinite distance from the front meniscus.
%\label{fig:heil:planar}}
%\end{figure}

@@

\subsection{The film variation over the bubble}
\label{section:film:thickness:variation}
Next, we investigated the variation of the film thickness over the bubble length.
The comparison for the thickness variation is presented in Fig.~\ref{fig:sehgal:bubble:length}.
\begin{figure}
\includegraphics[width=0.5\textwidth]{Figures/Bubble/bubble_sehgal_new.eps}\hfill
\includegraphics[width=0.5\textwidth]{Figures/Bubble/bubble_length_new.eps}\\
\caption{A qualitative comparison for the film thickness across
the bubble length. {\color{red} Left} (courtesy of \citet{sehgal-microchannel}),
{\color{red} Right}  (present simulations); $x$ is scaled to $H_{\mathrm{eff}}$
and increases in the flow direction.{\color{red} The data from
\cite{sehgal-microchannel} was digitally acquired by the program ``Engauge Digitizer'' and scaled
to compare with the present simulations.} \label{fig:sehgal:bubble:length}}
\end{figure}
One can see a qualitative agreement, i.e. the thickness increases towards the front meniscus and
rapidly decreases towards the rear meniscus. This shape is sometimes referred as a "bullet"
shape.

\subsection{Capillary number region}
\label{section:capillary:range}
The purpose of this section is to validate the correlations of
\citet{giavedoni-numerical} %in Fig. \ref{fig:giavedoni:planar} 
and
\citet{heil-bretherton} %in Fig. \ref{fig:heil:planar} 
for a range of capillary
numbers. Because of limited computational resources, we skip the
small capillary numbers and make calculations for the range of capillary numbers $0.03-0.8$, which
is a computationally reasonable task.  To be consistent with the previous sections, we choose the grid to be
$202 \times 3001$. Then $5$ lattice Boltzmann units do not occupy more than $60$
percent of the effective film thickness. Because the simulation gets unstable with
smaller grids and larger gradients, all the capillary number simulations were
performed on the same grid. To properly initialize the body force, the
proportionality law was utilized. The forcing
$6 \cdot 10^{-6}/16$ was chosen to obtain the predicted capillary
number $0.05$.  
The pressure gradient can be obtained through the capillary number
ratio, Section \ref{section:init}:
\frac{\mathrm{d}P}{\mathrm{d} x}=6 \cdot 10^{-6}/16 \frac{\Ca_{\mathrm{lit}}}{0.05}
The film thickness is initialized through the ratio of capillary numbers as well:
\begin{equation*}
12 \frac{\Ca_{\mathrm{lit}}}{0.05}
\end{equation*}
The results obtained after $2\cdot10^5$ steps are presented in Table
\ref{table:parameters:capillary:number}. 
{\color{red} We chose to measure the film thickness in the middle of the bubble. The
classical Bretherton formulation is for the measurement of the film thickness at ``infinity''.
However, \citet{cerro-bubble-train} indicate that for the gravity-driven bubble train flow in tubes
it is enough for bubble to have a length of two-three diameters to apply analysis for very long
bubbles. In comparison with the axisymmetric circular geometry, for three-dimensional square shaped
microchannels \citet{heil-threedim} indicate the measurement of the film thickness at $5.5$
channel heights from the front bubble tip. This distance was shown sufficient for the
film thickness to establish itself for the capillary number range $\Ca<4$.
\citet{giavedoni-numerical} showed that the film stabilizes at the distances as $2.6-4.0$ diameters
from the front tip depending on the Reynolds number.  We chose to measure the film thickness in the
middle of the bubble, which is located at least at the distance of $2.5-3$ channel heights from the
front bubble tip. Moreover, the film thickness examination, see Section
\ref{section:film:thickness:variation}, shows that for the capillary range of interest, i.e.
$0.05\leq \Ca \leq 1$, the standard deviation from the bubble middle film thickness is around $7$
percent for small capillary numbers ($\Ca=0.05$) and less than $1$ percent for larger ones ($\Ca
\geq 0.1$) at the distances ranging from one channel height from the rear tip upto one channel
height to the front tip. This certainly validates the point of choice to measure bubble thickness.}
One can see that the calculated
capillary numbers are overpredicted but the obtained thicknesses are overpredicted
as well due to the body force setting caused by the Poiseuille flow assumption. If one
needs to obtain the desired capillary number, the shooting method for the forcing is
necessary, starting with the Poiseuille pressure gradient force as the initial condition. We extracted
the data of 
\citet{giavedoni-numerical} and \citet{heil-bretherton} and compared them with our results (see 
Fig. \ref{fig:capillary:comparison}).
\includegraphics[width=0.97\textwidth]{Figures/Capillary/capillaries_comparison_real.eps}
\caption{A comparison between simulation results and results of
\citet{giavedoni-numerical} and \citet{heil-bretherton}. One can see a
reasonable agreement. The plots depict the film thickness as a function of the
the capillary number. {\color{red} Data from \cite{giavedoni-numerical} was
extracted with the help of ``Engauge Digital Discretizer software'', data from
\citet{heil-bretherton} was extracted by hands using a set of curves involving interpolation and
it's the reason for a data to exhibit a noisy trend.} \label{fig:capillary:comparison}}
